Estimating dark matter halo concentration using the integrated mass profile

### 2016-04-17 paper/paper.tex
new concentration plot
paper updated with new plots
updated plots and text

%*************************************************************************
\section{Introduction}
\label{sec:introduction}
%*************************************************************************
In the current structure formation paradigm the properties of galaxies
are coupled to the evolution of their dark matter (DM) hosting halo.
For instance, the sizes and dynamics of galaxies are determined by
the DM distribution inside a halo. This motivates the detailed study
a DM halo internal structure.

The internal DM distribution in a halo can be parameterized through the density
profile.  
In a first approximation this profile is spherically symmetric and
its density only depends on the radial coordinate. 
One the most popular radial parameterizations is the
Navarro-Frenk-White (NFW) profile 
\citep{NFW}.   
If one is not interested in the very central region of the
dark matter halo (where galaxy formation takes place, and where the
effects of baryon physics on dark matter distribution are still 
unknown) one can also assume that this profile is universal.
\citep{Navarro2010}.  
 This profile is a double power law in radius, where the transition
break happens at the so-called scale radius, $r_s$.  
The ratio between the scale radius and the halo virial radius $R_v$ is
known as the concentration $c=R_v/r_s$.

The concentration of the NFW profile provides the conceptual
framework to study simulated dark matter halos as a function of
redshift and cosmological parameters.
In many numerical studies 
\citep{Neto2007,Maccio2008,Duffy2008,Munoz2011,Prada2012,Ludlow2014} 
the results are summarized through the
mass-concentration relationship; that is the distribution of
concentration values at a fixed halo mass and redshift.
The success of such numerical experiments rests on a
reliable algorithm to estimate the concentration. 
Such an algorithm should provide unbiased results and must be robust
when applied to simulations of different numerical resolution.

There are two main algorithms to estimate the concentration parameter
of a N-body dark matter halo. 
The first method takes the halo particles and bins them into
logarithmic radii to estimate the density in each bin, then it 
proceeds to fit the density as a function of the radius to the NFW
profile.  
A second method uses an analytic property of the NFW profile
that relates the maximum of the ratio of the circular velocity to the
virial velocity, $V_{\rm circ}$/$V_{\rm vir}$.  The concentration can
be then found as the root of an algebraic equation dependent on this
maximum value.

@@

In this letter we present a third alternative based on fitting the
integrated mass profile.
This approach has two advantages with respect to the above mentioned
methods.  
It does not involve any data binning and does not throw away data
points. 
This translates into a robust estimate even at low resolution/particle
numbers.   
Furthermore, since the method does not require any binning, there is
no need to tune numerical parameters.
This provides a new independent method to estimate the concentration
parameter.   
Here we provide a detailed description of the method and a
comparison against the two traditional techniques by measuring its
performance both on mock and N-body halos.

@@

To date, there are two standard methods to estimate concentrations in
dark matter halos extracted from N-body simulations.

The first method tries to directly estimate the density profile.  
It takes all the particles in the halo and bins them in the logarithm of
the radial coordinate from the halo center.  
Then, it estimates the density in each logarithmic bin counting the
particles and dividing by the corresponding shell volume.  
At this point is possible to make a direct fit to the density as a
function of the radial coordinate. 
This method has been broadly used for more than two decades to study
the mass-concentration-redshift relation of dark matter halos.

@@

We apply the three methods mentioned above on two different halo
samples.
The first sample is composed by mock halos generated to have known
concentration values in perfect spherical symmetry following an NFW
profile.    
Using this sample we want to quantify the ability to recover the
expected concentration as a function of particle number.
The second sample comes from a publicly available N-body cosmological
simulation.  
In this way we test the new method on more realistic and complex
data. 
We summarize the results on this dataset by showing the impact on the
mass-concentration relationship.

\subsection{Mock halo sample}

@@

\label{sec:data}
\begin{figure*}
  \begin{center}
    \includegraphics[width=0.32\textwidth]{conc_mass_vel.pdf}
    \includegraphics[width=0.32\textwidth]{conc_mass_dens.pdf}
    \includegraphics[width=0.32\textwidth]{conc_dens_vel.pdf}
  \end{center}
\vspace{-0.5cm}
  \caption{Comparison between the concentrations measured by the new 
    method and the maximum velocity (left) and density (middle)
    methods. The solid line indicates the equal value between the two
    techniques. The dashed line shows the upper 15 per cent difference
    between two methods.
  \label{fig:comparison}}
\end{figure*}

@@

\begin{figure}
\begin{center}
  \includegraphics[width=0.50\textwidth]{concentration.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Mass-concentration relationship for the three different
  methods used on the same cosmological N-body data. 
  All halos are used regardless of their relaxation state.
  The lines correspond to the median concentration values in each mass bin.
  The shaded region presents 10 and 90 per cent spread.
  For clarity we only show the spread for the concentrations estimated
  using the new method. The other two methods have a similar spread. The
  dotted line corresponds to fits reported by \citep{Prada2012}.
  \label{fig:concentration}}
\end{figure}

@@

At fixed particle numbers the new method always shows the lowest
$\avg{|D|}$ values compared to the other two.
Its accuracy is on the order of $10\%$ for $20$ particles in the halo,
decreasing to $0.01\%$ for halos with $20000$ particles.  
The dependence of $\avg{|D|}$ with the particle number $N$
goes approximately as $\avg{|D|}\propto N^{-1/2}$.   
This hints that the increasing accuracy of the method could be due to
a decrease of Poisson noise.

@@

\subsection{Halos from a cosmological simulation}

@@

We use data from the MultiDark cosmological simulation that follows
the non-linear evolution of a dark matter density field sampled with
$2048^3$ particles over a cubic box of $1000\ \hMpc$ on a side.  
The data is publicly available at \url{http://www.cosmosim.org/}.
More details about the structure of the database and the simulation
can be found in \citep{2013AN....334..691R}.

We use a halo sample containing all the halos located in a cubic
sub-volume of $100$ \hMpc\ on a side centered on the most massive halo
in the simulation at $z=0$.
This dataset corresponds to the \texttt{miniMDR1}
table in the  database.  
From this sample we select all the halos at $z=0$ detected
with a Friends-of-Friends (FoF) algorithm with masses in the interval
$10^{12}\leq M_{\rm FoF}/\hMsun \leq 10^{14}$.  
The FoF algorithm used a linking length of $0.17$ times the mean
inter-particle distance. This choice translates into an overdensity
$\Delta_h\sim  400-700$ dependent on the halo concentration \citep{More2011}.
Finally, we select from the database all the particles that belong to
each FoF group.

From this set of particles we follow the procedure spelled out in
Section \ref{sec:method} with $\Delta_h=740$ (corresponding to $200$
times the critical density) to select an spherical region that we
redefine to be our halo.  
This choice makes that the overdensities are fully included inside the
original FoF particle group.   
On the interest of providing a fair comparison against the density method we
only report results from overdensities with at least $200$ particles.

Figure \ref{fig:comparison} shows the comparison between the
concentration values obtained by the three different methods. 
The diagonal line marks a one-to-one correspondence if all the values
were equal.
The three methods show a broad general agreement. 
The largest difference is present at high concentration $c>5$ where
the new method produces systematic lower values than the other two
methods. Most of the differences are at the $20\%$ level, with some
extreme cases for the highest concentration values with $50\%$
differences.  The velocity and density methods produce consistent
values within a $20\%$ scatter, only some halos with very high
concentrations $c\sim 10$ have a $15\%$ lower concentration with the
velocity method.

We explored halo shape and relaxation as a source of this
differences. We found that shape does not bias the concentration
estimation; relaxation, quantified by the
offset between the center of mass and minimum of the potential, can
only offset at most $5\%$ the median results between methods.  
We note that the highest concentration values correspond to the smallest halos
with $\sim 200$ particles, a regime where both the density and
velocity methods are reliable only the $5\%$ in a perfect spherical
halo.

The differences we find might signal a deviation from the NFW profile
due to the low resolution of these halos. We suggest that asserting
with confidence the systematic difference between the different
methods should be confirmed  a higher resolution simulation and larger
statistics. We leave this comparison for future work.

Figure \ref{fig:concentration} summarizes the previous results in the
mass-concentration relationship. 
We find that the median concentration
can be fit by the following relationship
\begin{equation} c_{200} = 5.62 \left(\frac{M_{200}}{10^{12}\hMsun}\right)^{-0.061}, 
\end{equation}
%
which has a similar power value compared to the $-0.075$ reported by
\citep{Prada2012}, while the normalization is $\sim 20\%$ lower
compared to the $7.28$ value reported by the same authors.

>>>>

%*************************************************************************
\section{Introduction}
\label{sec:introduction}
%*************************************************************************
In the current structure formation paradigm the properties of galaxies
are coupled to the evolution of their dark matter (DM) hosting halo.
In this paradigm the sizes and dynamics of galaxies are driven by
the DM distribution inside a halo. This has motivated the detailed study
a DM halo internal structure in the last three decades.

The internal DM distribution in a halo is usually parameterized
through the density profile.  
In a first approximation this profile is spherically symmetric; the
density only depends on the radial coordinate.  
One the most popular radial parameterizations is the
Navarro-Frenk-White (NFW) profile 
\citep{NFW}.   
If one is not interested in the very central region of the
dark matter halo (where galaxy formation takes place, and where the
effects of baryon physics on dark matter distribution are still 
unknown) one can also assume that this profile is universal.
\citep{Navarro2010}.  
 This profile is a double power law in radius, where the transition
break happens at the so-called scale radius, $r_s$.  
The ratio between the scale radius and the halo virial radius $R_v$ is
known as the concentration $c=R_v/r_s$.

The concentration of the NFW profile provides a conceptual
framework to study simulated dark matter halos as a function of
redshift and cosmological parameters.
In many numerical studies 
\citep{Neto2007,Maccio2008,Duffy2008,Munoz2011,Prada2012,Ludlow2014} 
the results are summarized through the
mass-concentration relationship; that is the distribution of
concentration values at a fixed halo mass and redshift.
The success of such numerical experiments rests on a
reliable algorithm to estimate the concentration. 
Such an algorithm should provide unbiased results and must be robust
when applied to simulations of different numerical resolution.

There are two stablished algorithms to estimate the concentration parameter
of a N-body dark matter halo. 
The first method takes the halo particles and bins them into
logarithmic radii to estimate the density in each bin, then it 
proceeds to fit the density as a function of the radius to the NFW
profile.  
A second method uses an analytic property of the NFW profile
that relates the maximum of the ratio of the circular velocity to the
virial velocity, $V_{\rm circ}$/$V_{\rm vir}$.  The concentration can
be then found as the root of an algebraic equation dependent on this
maximum value.

@@

In this letter we present a third alternative based on fitting the
integrated mass profile.
This approach has two advantages with respect to the above mentioned
methods.  
It does not involve any data binning and does not throw away data
points. 
This translates into a robust estimate even at low resolution/particle
numbers.   
Furthermore, since the method does not require any binning, there is
no need to tune numerical parameters.
This method provides a new independent method to estimate the
concentration parameter.   
Here we provide a detailed description of the method and a
comparison against the two traditional techniques by measuring its
performance both on mock and N-body halos.

@@

To date, there are two standard methods to estimate concentrations in
dark matter halos extracted from N-body simulations. 
The first method tries to directly estimate the density profile.  
It takes all the particles in the halo and bins them in the logarithm of
the radial coordinate from the halo center.  
Then, it estimates the density in each logarithmic bin counting the
particles and dividing by the corresponding shell volume.  
At this point is possible to make a direct fit to the density as a
function of the radial coordinate. 
This method has been broadly used for more than two decades to study
the mass-concentration-redshift relation of dark matter halos.

@@

We apply the three methods mentioned above on two different halo
samples.
The first sample is composed by mock halos generated to have known
concentration values in perfect spherical symmetry following an NFW
profile.    
Using this sample we quantify the ability to recover the expected
concentration as a function of particle number. 
The second sample comes from a publicly available N-body cosmological
simulation to test the new method on more realistic and complex data
to show the impact on the mass-concentration relationship.

\subsection{Retrieving known concentrations from a mock halo sample}

@@

%\label{sec:data}
%\begin{figure*}
%  \begin{center}
%    \includegraphics[width=0.32\textwidth]{conc_mass_vel.pdf}
%    \includegraphics[width=0.32\textwidth]{conc_mass_dens.pdf}
%    \includegraphics[width=0.32\textwidth]{conc_dens_vel.pdf}
%  \end{center}
%\vspace{-0.5cm}
%  \caption{Comparison between the concentrations measured by the new 
%    method and the maximum velocity (left) and density (middle)
%    methods. The solid line indicates the equal value between the two
%    techniques. The dashed line shows the upper 15 per cent difference
%    between two methods.
%  \label{fig:comparison}}
%\end{figure*}

@@

@@

At fixed particle numbers the new method always shows the lowest
$\avg{|D|}$ values compared to the other two.
Its accuracy is on the order of $10\%$ for $20$ particles in the halo,
decreasing to $0.01\%$ for halos with $20000$ particles.  
The dependence of $\avg{|D|}$ with the particle number $N$
goes approximately as $\avg{|D|}\propto N^{-1}$.

@@

We also find that the bias from different concentrations at fixed
particle number is negligeable with respect to the effects induced by
different particle numbers.

\begin{figure}
\begin{center}
  \includegraphics[width=0.47\textwidth]{avg_foff_bolshoi.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Average value of the relative error in the concentration
  estimate, $\avg{|D|}$, as a function of the particle number $N$ in
  the set of mock halos. Different symbols represent different
  methods. The new method provides the most accurate estimate at fixed
  particle number.  
    \label{fig:downsampling}}
\end{figure}

\begin{figure}
\begin{center}
  \includegraphics[width=0.45\textwidth]{error_conc_bolshoi.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Average value of the relative error in the concentration
  estimate, $\avg{|D|}$, as a function of the particle number $N$ in
  the set of mock halos. Different symbols represent different
  methods. The new method provides the most accurate estimate at fixed
  particle number.  
    \label{fig:downsampling}}
\end{figure}

\subsection{Measuring the concentration in halos from a cosmological simulation}

\begin{figure*}
\begin{center}
  \includegraphics[width=0.45\textwidth]{concentration_bolshoi.pdf}
  \includegraphics[width=0.45\textwidth]{concentration_bolshoi_corrected.pdf}
\end{center}
\vspace{-0.5cm}
\caption{Mass-concentration relationship for the three different
  methods used on the same cosmological N-body data. 
  All halos are used regardless of their relaxation state.
  The lines correspond to the median concentration values in each mass bin.
  The shaded region presents 10 and 90 per cent spread.
  For clarity we only show the spread for the concentrations estimated
  using the new method. The other two methods have a similar spread. The
  dotted line corresponds to fits reported by \citep{Prada2012}.
  \label{fig:concentration}}
\end{figure*}

@@

We use data from the Bolshoi cosmological simulation that follows
the non-linear evolution of a dark matter density field sampled with
$2048^3$ particles over a cubic box of $250\ \hMpc$ on a side.  
The data is publicly available at \url{http://www.cosmosim.org/}.
More details about the structure of the database and the simulation
can be found in \citep{2013AN....334..691R}.

We use a halo sample containing all the halos located in a cubic
sub-volume of $100$ \hMpc\ on a side.
From this sample we select all the halos at $z=0$ detected
with a Friends-of-Friends (FoF) algorithm with more than 300
particles, meaning that the masses are in the interval
$4\times 10^{10}\leq M_{\rm FoF}/\hMsun \leq 10^{14}$.  
The FoF algorithm used a linking length of $0.17$ times the mean
inter-particle distance. This choice translates into an overdensity
$\Delta_h\sim  400-700$ dependent on the halo concentration \citep{More2011}.

From this set of particles we follow the procedure spelled out in
Section \ref{sec:method} with $\Delta_h=740$ (corresponding to $200$
times the critical density) to select an spherical region that we
redefine to be our halo.  
This choice makes that the overdensities are fully included inside the
original FoF particle group.   
On the interest of providing a fair comparison against the density method we
only report results from overdensities with at least $200$ particles
($2.6\times 10^{10}$\hMsun).

\subsubsection{Convergence Test using Downsampling}

We perform first a convergence test on the N-body data to find
possible biases in the concentration estimates as a function of
particle number. 
We select $8$ halos from the simulation that have at least $3\times
10^5$ particles. 
We then downsample the total number of particles in each halo by
factors of $10, 20, 45, 100, 215, 460, 1000$ to measure its
concentration with both the integrated mass and the velocity method.
each downsampling factor we repeat the operation $100$ times.

The average concentration value for the largest number of particles, $c_{N_{max}}$,
provides with a baseline to compare all the other results.
We use the following statistic
\begin{equation}
f_{\rm off} = c_{N} / c_{N_{max}} - 1,
\end{equation}
to account for the offset between the concentration at a given
downsampled particle number $c_{N}$ and the baseline $c_{N_{max}}$.

Figure \ref{fig:downsampling} summarizes our results. 
The plot shows the average value of $f_{\rm off}$ as a function
of the particle number.
As expected for large enough sample particle numbers, $N_{s}>5\times
10^3$, the results of the two algorithms coincide and $\langle f_{\rm
  off}\rangle\sim 0$.  
For lower number of particles, the results of the two algorithms
deviate from the expected concentration value.  
The velocity method overestimates by a factor $\langle f_{\rm 
  off}\rangle\sim 0.25$ the concentration values for the lowest
sample particle numbers, $N_{s}\sim 100$. 
Around the same sampling scale, the new algorithm shows a more stable
behaviour with an understimate by a factor of $\langle f_{\rm 
  off}\rangle \sim 0.02$.  
This shows that in the more complex setup of a cosmological setup our
algorithm continues to be stable below the $5\%$ level for low 
particle numbers.

The line on the same panel shows a fit to the data with the following
functional form 
\begin{equation}
f_{\rm off} = \frac{A}{(1+\log_{10}N_s)^{B}}.
\end{equation} 
for the dataset shown in the panel we have $A=3385\pm 2065$, $B=7.99\pm0.48
$ for the velocity method and $A=-110\pm 253$, $B=7.11\pm1.80$ for the
new method, where the uncertainties are estimated from the covariance 
matrix resulting from the non-linear optimization procedure. The lines
on Figure \ref{fig:downsampling} are plotted using these best values.

Figure \ref{fig:downsampling} summarizes different uncertainty
results. 
Large symbols show the standard deviation on $f_{\rm off}$
from the downsampling experiment.
Small dots show the uncertainty on each individual concentration value
as estimated by the MCMC algorithm.
The line shows the $\propto N^{-1}$  locus.
The first thing to notice is that statistical uncertainties are at
least one order of magnitude than the uncertainty associated to
downsampling experiment.

\subsubsection{Mass-Concentration Relationship}

Figure \ref{fig:concentration} summarizes the previous results in the
mass-concentration relationship. 
We find that the median concentration
can be fit by the following relationship
%\begin{equation} c_{200} = 5.62 \left(\frac{M_{200}}{10^{12}\hMsun}\right)^{-0.061}, 
%\end{equation}
%
which has a similar power value compared to the $-0.075$ reported by
\citep{Prada2012}, while the normalization is $\sim 20\%$ lower
compared to the $7.28$ value reported by the same authors.
